New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-09-12 paper/appendix.Rmd
went through sections 1 and 2 and moved a paragraph of mathematica to appendix as requested by Ursula

Other functions are:

>>>>

The Slice Tour* notebooks include the additional functions: ProjectionPlot, Projected2DSliderPlot, and VisualiseSliceDyanmic. ProjectionPlot is very similar to SliceDynamic, except it only displays projections and VisualiseSliceDynamic allows you to discern which points exist inside the slice and outside it. However, ProjectedLocatorPlot displays the interactive slightly differently. Several locator panes are displayed above the plot and each one corresponds to a row in the projection matrix but behaves the same as ProjectionPlot.

### 2022-09-12 paper/paper.Rmd
went through sections 1 and 2 and moved a paragraph of mathematica to appendix as requested by Ursula

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the coefficients of a variable in a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the slice center, which is useful for exploring how structure changes in local subspaces. These controls are implemented in Mathematica to take advantage of it's mathematics functions, and interactivity tools that allow construction of functions that interact with the position of the cursor. We demonstrate the usefulness of this new tool with an example exploring region boundaries in classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approch is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many, many tour developments, which are summarized in @lee2021.

@@

<!-- Purpose -->
Manual controls are especially useful for assessing sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, can some variables be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to control the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the value of center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center.

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods are illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a mathematica notebook that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and two mathematica notebooks, one focused on the application, and the other that could be used generically for new data.

@@

The method for updating component values in @cook_manual_1997 (and utilised in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

@@

The problem with new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. There are numerous ways that this can be enforced, and details are in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

@@

Mathematica provides much utility and versatility, such as inbuilt data visualisation, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is quite user-friendly and can be found in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there. Also, it is relatively simple  to create dynamic objects via the inbuilt commands `Manipulate` or `DynamicModule`. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that dynamic variable will be updated. This is the essential ingredient for this implementation of the manual tour. However, it can be hard to alter the behaviour of the control objects such as setting the 1D slider on a different scale  (Slider increments add 10 points into the slice rather than increasing the slice height by 0.1 for example).
{\textcolor{red}Alex, is this still true with the controls you are using now?}

The examples presented in the penguins’ notebooks all utilise our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index. After specifying the initial slice height and the slice range, the user is then presented with an interactive display in which the control objects appear on the left and the slice visulisation appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice height; and there is an input field, which changes the slice centre. The user can also change the appearance of the plot by zooming into the center or changing the point size with sliders provided. The current projection can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can be displayed via another checkbox. The Slice Tour* notebooks include the additional functions: ProjectionPlot, Projected2DSliderPlot, and VisualiseSliceDyanmic. ProjectionPlot is very similar to SliceDynamic, except it only displays projections and VisualiseSliceDynamic allows you to discern which points exist inside the slice and outside it. However, ProjectedLocatorPlot displays the interactive slightly differently. Several locator panes are displayed above the plot and each one corresponds to a row in the projection matrix but behaves the same as ProjectionPlot. {\textcolor{blue}(this may be easier to implement in another language (?) )}

>>>>

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the \textcolor{blue} {coefficients of a variable in} \textcolor{red}{contribution of a given variable to} a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the slice center, which is useful for exploring how structure changes in local subspaces. These controls are implemented in Mathematica to take advantage of \textcolor{blue}{it's} \textcolor{red}{its} \textcolor{blue}{remove this: mathematics functions, and interactivity tools} \textcolor{red}{location controls} that allow construction of functions that interact with the position of the cursor. We demonstrate the usefulness of this new tool with an example exploring region boundaries in classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approch is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many \textcolor{blue}{remove: }, many} tour developments, which are summarized in @lee2021.

@@

<!-- Purpose -->
Manual controls are especially useful for assessing sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, \textcolor{red}{} one may ask whether some} variables can be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to \textcolor{blue}{control}\textcolor{red}{choose} the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the value of center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. \textcolor{red}{It would also allow the user to interactively decide on the thickness of the slice.}

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods \textcolor{blue}{are}\textcolor{red}{ is} illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a mathematica notebook that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and two mathematica notebooks, one focused on the application, and another that could be used generically for new data.

@@

The methods for updating component values in @cook_manual_1997 (and utilised in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

@@

The problem with \textcolor{red}{the} new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. \textcolor{red}{Even though these changes are for the most part imperceptible, one may wits to remove them and} there are numerous ways that this can be enforced, and details are given in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

@@

Mathematica provides much utility and versatility, such as inbuilt data visualisation, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is quite user-friendly and can be found in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there. Also, it is relatively simple  to create dynamic objects via the inbuilt commands `Manipulate` or `DynamicModule`. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that dynamic variable will be updated. This is the essential ingredient for this implementation of the manual tour. However, it can be hard to alter the behaviour of the control objects such as setting the 1D slider on a different scale  (Slider increments add 10 points into the slice rather than increasing the slice height by 0.1 for example). \textcolor{red}{but this is no longer true if you replace it with animate??}
\textcolor{red}{Alex, is this still true with the controls you are using now?}

The examples presented in the penguins’ notebooks all utilise our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index \textcolor{red}{(which can be one if there is only one group)}. After specifying the initial slice thickness and the slice range, the user is then presented with an interactive display in which the control objects appear on the left and the slice visulisation appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice centre. The user can also change the appearance of the plot by zooming into the center or changing the point size with sliders provided.\textcolor{red}{this zoom doesn't work very well} The current projection can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox in order to store it for further study.

### 2022-09-12 paper/paper.tex
went through sections 1 and 2 and moved a paragraph of mathematica to appendix as requested by Ursula

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the coefficients of a variable in a low-dimensional
projection, which is useful for exploring the sensitivity of structure
to particular variables. The user can also interactively shift the slice
center, which is useful for exploring how structure changes in local
subspaces. These controls are implemented in Mathematica to take
advantage of it's mathematics functions, and interactivity tools that
allow construction of functions that interact with the position of the
cursor. We demonstrate the usefulness of this new tool with an example
exploring region boundaries in classification problems.
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many, many tour developments, which are
summarized in \citet{lee2021}.

@@

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, can some variables be removed from the
projection without affecting the clustering. For interpreting models,
one can reduce or increase a variable's contribution to examine the
variable importance. Having the user interact with a projection is
extremely valuable for understanding high-dimensional data. However,
these algorithms have two problems: (1) the pre-processing of creating a
manipulation space overly complicates the algorithm, (2) extending to
higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to
control the position of the center of a slice. The slice tour was
introduced in \citet{slicetour}. It operates by converting the
projection plane into a slice, by removing or de-emphasizing points that
are further than a fixed orthogonal distance from the plane. The
projection plane is usually thought of as passing through the center of
the data. Manual control would allow the user to change the value of
center point, by shifting it along a coordinate axis, while keeping the
orientation of the projection plane fixed. The purpose would be to
explore how or if the shape of the data, in the space orthogonal to the
projection, changes as one gets away from the center.

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods are
illustrated to compare and contrast boundaries constructed by different
classifiers. The software section describes a mathematica notebook that
is used for the application, and describes the interactive environment
that would be desirable within R as new technology becomes available.
The paper is accompanied by an appendix with more details and
adjustments to the manual controls, and two mathematica notebooks, one
focused on the application, and the other that could be used generically
for new data.

@@

The method for updating component values in \citet{cook_manual_1997}
(and utilised in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The problem with new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. There are numerous ways that this can be enforced, and details are
in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

>>>>

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the
\textcolor{blue} {coefficients of a variable in}
\textcolor{red}{contribution of a given variable to} a low-dimensional
projection, which is useful for exploring the sensitivity of structure
to particular variables. The user can also interactively shift the slice
center, which is useful for exploring how structure changes in local
subspaces. These controls are implemented in Mathematica to take
advantage of \textcolor{blue}{it's} \textcolor{red}{its}
\textcolor{blue}{remove this: mathematics functions, and interactivity tools}
\textcolor{red}{location controls} that allow construction of functions
that interact with the position of the cursor. We demonstrate the
usefulness of this new tool with an example exploring region boundaries
in classification problems.
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many \textcolor{blue}{remove: }, many\}
tour developments, which are summarized in \citet{lee2021}.

@@

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, \textcolor{red}{} one may ask whether some\}
variables can be removed from the projection without affecting the
clustering. For interpreting models, one can reduce or increase a
variable's contribution to examine the variable importance. Having the
user interact with a projection is extremely valuable for understanding
high-dimensional data. However, these algorithms have two problems: (1)
the pre-processing of creating a manipulation space overly complicates
the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to
\textcolor{blue}{control}\textcolor{red}{choose} the position of the
center of a slice. The slice tour was introduced in \citet{slicetour}.
It operates by converting the projection plane into a slice, by removing
or de-emphasizing points that are further than a fixed orthogonal
distance from the plane. The projection plane is usually thought of as
passing through the center of the data. Manual control would allow the
user to change the value of center point, by shifting it along a
coordinate axis, while keeping the orientation of the projection plane
fixed. The purpose would be to explore how or if the shape of the data,
in the space orthogonal to the projection, changes as one gets away from
the center.
\textcolor{red}{It would also allow the user to interactively decide on the thickness of the slice.}

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods
\textcolor{blue}{are}\textcolor{red}{ is} illustrated to compare and
contrast boundaries constructed by different classifiers. The software
section describes a mathematica notebook that is used for the
application, and describes the interactive environment that would be
desirable within R as new technology becomes available. The paper is
accompanied by an appendix with more details and adjustments to the
manual controls, and two mathematica notebooks, one focused on the
application, and another that could be used generically for new data.

@@

The methods for updating component values in \citet{cook_manual_1997}
(and utilised in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The problem with \textcolor{red}{the} new simple method is that it is
not faithful to the precise values for \(V_m\) because the
orthonormalisation will change them.
\textcolor{red}{Even though these changes are for the most part imperceptible, one may wits to remove them and}
there are numerous ways that this can be enforced, and details are given
in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}
